\large How resource abundance and stochasticity affect animals' spatial needs

### 2022-11-18 writing/manuscript.Rmd
end of week push

---
title: \large How resource abundance and stochasticity affect animals' spatial needs
author:
  - name: Stefano Mezzini
    email: stefano.mezzini@ubc.ca
    institute: ubco
  - name: Emilia Patrícia Medici
    institute: [tapir, iucn, escas]
  - name: Michael J. Noonan
    email: michael.noonan@ubc.ca
    institute: ubco
institute:
  - ubco: The Irving K. Barber Faculty of Science, The University of British Columbia, Okanagan Campus, Kelowna, Canada.
  - tapir: Lowland Tapir Conservation Initiative (LTCI), Instituto de Pesquisas Ecológicas (IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960‑000, Brazil.
  - iucn: IUCN SSC Tapir Specialist Group (TSG), Campo Grande, Brazil.
  - escas: Escola Superior de Conservação Ambiental E Sustentabilidade (ESCAS/IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960‑000, Brazil.
bibliography: 'env-var-review.bib'
csl: 'the-american-naturalist.csl'
fontsize: 12pt
indent: true
header-includes:
    - \usepackage{setspace}\doublespacing # for double-spaced text
    - \usepackage{indentfirst} # for indenting first line of each paragraph
    - \usepackage[small]{titlesec} # for smaller font for headings
    - \usepackage{wrapfig} # for wrapping text around figures
    - \usepackage{caption} # for more customization of captions
    # smaller caption text, less space above and below caption
    - \captionsetup[figure]{font=scriptsize, aboveskip=4pt, belowskip=-15pt}
    - \usepackage{afterpage} # for \afterpage{\<command>}, e.g. \afterpage{\clearpage}
    # - \usepackage{gensymb} # for \degree, \celsius, \perthousand, \micro and \ohm
    # - \usepackage{float} # for better floating objects interface; adds more floats 
    # - \usepackage{ragged2e} # for \Centering, \RaggedLeft, and \RaggedRight
subparagraph: true # needed for \usepackage[small]{titlesec}
linenumbers: true
output:
  bookdown::pdf_document2:
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
    toc: false
    number_sections: false
---

@@

***Not sure if we need this or if we can tie it in somewhere:***

@@

\noindent While animals' needs vary greatly between taxonomic groups, some needs are essential for most species for survival and reproduction. All heterotrophic organisms require sources of chemical energy (i.e., food), water, and various limiting nutrients to survive, grow, and reproduce [@harvey_primate_1981; @baldwin_nutritional_1984; @reich_body_2001]. Failing to acquire sufficient resources is likely to result in lower fitness, changes in physiology or behavior, physiological damage, inability to reproduce, and death. Motile organisms can move to new locations in search of resources, but they must also account for a higher metabolism and movement costs [@harvey_primate_1981; @ref]. The size of an animal's spatial needs is often hypothesized to be depend on resource abundance [@burt_territoriality_1943], such that spatial needs increase when resources are low, but the relationship is likely not linear [see figure \@ref(fig:mu-hyp) and @nilsen_can_2005; @herfindal_prey_2005; @bista_effect_2022]. Low values of $\e(R)$ force animals to have larger home ranges [@relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005; @bista_effect_2022], migrate to better locations [@samarra_movements_2017; @middleton_green-wave_2018; @geremia_migrating_2019], or move nomadically [@singh_migration_2012; @polansky_elucidating_2015; @teitelbaum_beyond_2019; @nandintsetseg_variability_2019], although large-scale changes in behavior (such as species-wide shifts to dispersal, migration, and nomadism) are more likely to occur over evolutionary timescales rather than an animal's lifespan (figure \@ref(fig:mu-hyp)). As $\e(R)$ increases, we expect animals' spatial requirements to decrease nonlinearly until they reach the smallest amount of space required to survive [see @relyea_home_2000; @bista_effect_2022, although the latter uses models which assumes the effect to be linear].

@@

\noindent Consider the simplistic scenario where the average in resource abundance, $\e(R) = \mu$, changes over time but $\var(R) = \sigma^2$ is constant over time and space (and non-zero, see figure \@ref(fig:multipanel-constant-var)a). We can use the notation $R \sim \Gamma\big(\mu(t), \sigma^2\big)$ to indicate that $R$ follows a Gamma distribution with a mean $\mu(t)$ that changes over time (i.e., it is a function of time) while the variance is constant (figure \@ref(fig:multipanel-constant-var)b). Since $R$ is spatiotemporally random, an animal that moves in the landscape will not find the same $R$ at different time points or throughout the habitat (figures \@ref(fig:multipanel-constant-var)c-d). However, if $\mu(t)$ changes repetitively and regularly over time (e.g. peaks each spring and decreases in winter), an animal may learn to predict times of high or low $R$ [e.g., @falcon-cortes_hierarchical_2021; @geremia_migrating_2019; @abrahms_memory_2019; @samarra_movements_2017; @abrahms_memory_2019] although it will not be possible to predict where high-$R$ locations are, since $\e(R)$ does not vary predictably over space.

@@

@@

Environmental variability, including extreme events [@logares_black_2012], can reduce a landscape's energetic balance [@chevin_adaptation_2010], which, in turn, decreases animals' fitness and increases their spatial needs. A few recent studies support this hypothesis [@morellet_seasonality_2013; @nandintsetseg_variability_2019; @riotte-lambert_environmental_2020], but many of them are limited in geographic and taxonomic scales, so the extent to which these preliminary findings can be generalized is still very limited. There thus remains a need for developing a more complete understanding of how animals' spatial needs change with environmental stochasticity.

@@

Most habitats with appreciable seasonal changes in climate will require models to account for changes in both $\e(R)$ and $\var(R)$, since periods of high productivity can often be more variable than times of dormancy or lower productivity, and the variability in $R$ can often change both within and between years. In highly variable and stochastic environments, animals will likely have higher spatial needs, particularly if they are required to collect and store resources for periods of scarcity or increased energetic needs. These include reproductive periods, whether this be due to pregnancy [@ref], to feed offspring [@ref], or because the animal's needs change over time [e.g., male deer growing antlers: @french_nutrient_1956; @smolko_soil_2022]. Thus, failing to account for changes in $\var(R)$ can result in serious biases in the estimation of animals' spatial needs [@examples].

@@

Not all 25 scenarios depicted in figure \@ref(fig:5-5-sims) may be realistic, but the trends in $\e(R)$ and $\var(R)$, and their impacts on animal space use are useful examples that can be thought of as simplified scenarios. $\e(R)$ and $\var(R)$ can be assumed to be (approximately) constant in highly homogeneous environments, or environments where resources are sufficiently abundant that changes in $\e(R)$ and $\var(R)$ are relatively small and remain undetected. Although it is impossible for $\e(R)$ and $\var(R)$ to increase linearly forever, such increases may be possible for short periods of time (followed by periods of no change or decrease). Cyclical oscillations in $\e(R)$ and $\var(R)$ may occur in urban environments [@peron_periodic_2017; @ikeda_home_2022] and as temperatures fluctuate daily and seasonally [@geremia_migrating_2019], while $\e(R)$ and $\var(R)$ may drift randomly in highly complex environments which are too hard to predict. Finally, erratic changes in $\e(R)$ and $\var(R)$ may occur in environments where changes are very sudden, such as areas prone to fires or floods, as well as habitats with drastic human alteration (e.g., a forest which is clear-cut for mining purposes with a subsequent artificial re-forestation). However, if changes are stochastic but sufficiently small and frequent, animals may perceive them as few smooth changes rather a series of small, sudden, changes. Estimating trends in $\e(R)$ and $\var(R)$ requires a high cognitive capacity, since distinguishing between $\var (R)$ and changes in $\e(R)$ is not easy [@steixner-kumar_strategies_2020], especially if $\e(R)$ and $\var(R)$ are not independent (as in the case of $R \sim \Gamma(\mu(t), \sigma^2(t))$).

Although the trends in figure \@ref(fig:5-5-sims) are complex and the effects of $\e(R)$ and $\var(R)$ do not seem easy to disentangle, the effects of $\e(R)$ and $\var(R)$ are as we hypothesized (figure \@ref(fig:5-5-reg)). In the following discussion section, we detail how this framework can be applied to empirical data and used for conservation decision-making.

@@

This paper presents a new unifying framework for estimating the effects of resource abundance and stochasticity on animals' spatial needs. The hypotheses and methods we present allow researchers to build upon previous findings and datasets rather than stating their results as *de novo* findings. Instead, one can test whether members of the same population or species behave similarly, and whether all members of the group have a common functional response [@ref]. Individuals and populations that were once said to have markedly different behaviors and spatial needs may instead have a single common behavior but with different levels of $\e(R)$ and $\var(R)$ [@examples].

In this discussion section, we make considerations about the effects of resource abundance and stochasticity on various adaptations, including movement behaviors, energy storage, and spatiotemporal memory. We then illustrate how one may apply this framework to their own data and we also provide an empirical example using tracking data on a lowland tapir (*Tapirus terrestris*) from the Brazilian Cerrado. Finally, we demonstrate how to use this framework to create products that can inform conservation-based decisions, including predicting animals' spatial needs and assessing the quality of potential (and current) conservation areas.

* failing to account for $\var(R)$ may lead to decreases in population fitness and size as well as increase the risk of population collapse or human-wildlife conflict
* find examples of extreme events that caused population collapses or mass die-offs
* increasing stability of ecosystems and populations is a good way of increasing chances of long-term survival and fitness

## The spatiotemporal scale of stochastic events: memory and adaptability

@ranc_memory_2022: Roe deer create HR based on memory

The spatiotemporal scales over which an event occurs are a main determinant of whether an animal will be able to predict the event and how it will respond to it (or fail to do so). Events which occur at the scale of the organism (i.e. approximately one or more times per lifetime and within the animal's spatial range) are more likely to be perceived as a threat or boon worth preparing for [e.g., elephants: @foley_severe_2008; bison and mule deer: @geremia_migrating_2019]. Smaller organisms (e.g., mice) are more likely to be be severely impacted by a stochastic event than larger ones (e.g. elephants), since larger organisms can have bigger energy reserves [@lindstedt_seasonality_1985], can move longer distances over short periods of time [@hirt_general_2017], and tend to have longer lifespans, generation times, and developmental periods [@brown_toward_2004] that allow them to develop or memory about the frequency and severity of such events [@foley_severe_2008; @polansky_elucidating_2015]. However, the short generation time and high reproductive rate of smaller *r-selected* species [@pianka_r-_1970; @brown_toward_2004] may also promote traits that increase survival following frequent extreme events [@examples]. Frequent events may result in changes in animal behavior and memory [framework: @riotte-lambert_environmental_2020; elephants: @foley_severe_2008; bison and mule deer: @geremia_migrating_2019; elk: @falcon-cortes_hierarchical_2021], while events that occur over larger spatiotemporal scales are more likely to promote slower changes through adaptation, natural selection, and evolution [@logares_black_2012; @examples], including the developing of memory [framework: @riotte-lambert_environmental_2020; **mule deer?**: @ranc_memory_2022; elk: @falcon-cortes_hierarchical_2021; blue whales: @abrahms_memory_2019; elephants: @polansky_elucidating_2015; **species?**: @fagan_spatial_2013].

Adaptability and responsiveness to change are strong determinants to whether an animal survives in a changing environment [@riotte-lambert_environmental_2020]. An animal's ability to recognize cues and anticipate periods and areas of scarcity or abundance can have appreciable effects on the animal's chance of survival and reproduction. @rickbeil_plasticity_2019 showed that the yearly migration of elk (*Cervus canadensis*) depends on cues such as available forage biomass, hunting pressure, snow fall, and snow melt, while @falcon-cortes_hierarchical_2021 propose that elk move following nonrandom patterns based on memory that can last longer than 11 months. @jonzen_rapid_2006 demonstrated that large-scale, trans-Saharian migrant birds adapt their migration timing based on changes in the beginning of spring. However, relying strongly on short-term cues rather than depending on memory can also cause animals to fail to find crucial resources [@polansky_elucidating_2015; @foley_severe_2008] or become trapped in sub-optimal locations if long-term conditions and cycles change unexpectedly [e.g., blue whales: @abrahms_memory_2019; bison: @merkle_spatial_2019]. *changes in bird migration, check for changes in mammals*

@@

* The hypotheses we present here allow researchers to combine findings and generalize
* the analytical methods we use provide an example for how researchers can test these hypotheses using their own data
* can present results building on previous work rather than listing them as *de novo* findings
* additionally, can test for common behaviors between members of the same population or species
* ecologists often conclude that populations from different regions behave differently [@examples], but their differences in behavior may only be due to differences in $\e(r)$ or $\var(R)$
* using the methods presented here, it is possible to test whether populations and individuals do behave differently while also quantifying the differences in behavior

@@

@@

@@

## Lowland tapir in the Brazilian Cerrado: an empirical example

@@

## Using this framework in conservation-based decisions

>>>>

---
title: \large How resource abundance and stochasticity affect animals' spatial needs
author:
  - name: Stefano Mezzini
    email: stefano.mezzini@ubc.ca
    institute: ubco
  - name: Emilia Patrícia Medici
    institute: [tapir, iucn, escas]
  - name: Michael J. Noonan
    email: michael.noonan@ubc.ca
    institute: ubco
institute:
  - ubco: The Irving K. Barber Faculty of Science, The University of British Columbia, Okanagan Campus, Kelowna, Canada.
  - tapir: Lowland Tapir Conservation Initiative (LTCI), Instituto de Pesquisas Ecológicas (IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960‑000, Brazil.
  - iucn: IUCN SSC Tapir Specialist Group (TSG), Campo Grande, Brazil.
  - escas: Escola Superior de Conservação Ambiental E Sustentabilidade (ESCAS/IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960‑000, Brazil.
bibliography: 'env-var-review.bib'
csl: 'the-american-naturalist.csl'
fontsize: 12pt
indent: true
header-includes:
    - \usepackage{setspace}\doublespacing # for double-spaced text
    - \usepackage{indentfirst} # for indenting first line of each paragraph
    - \usepackage[small]{titlesec} # for smaller font for headings
    - \usepackage{wrapfig} # for wrapping text around figures
    - \usepackage{caption} # for more customization of captions
    # smaller caption text, less space above and below caption
    - \captionsetup[figure]{font=scriptsize, aboveskip=4pt, belowskip=-15pt}
    - \usepackage{afterpage} # for \afterpage{\<command>}, e.g. \afterpage{\clearpage}
    # - \usepackage{gensymb} # for \degree, \celsius, \perthousand, \micro and \ohm
    # - \usepackage{float} # for better floating objects interface; adds more floats 
    # - \usepackage{ragged2e} # for \Centering, \RaggedLeft, and \RaggedRight
subparagraph: true # needed for \usepackage[small]{titlesec}
linenumbers: true
output:
  bookdown::pdf_document2:
    latex_engine: xelatex # @reed_frequency_2003 has unicode character U+0308
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
    toc: false
    number_sections: false
---

@@

@@

\noindent While animals' needs vary greatly between taxonomic groups, some needs are essential for most species for survival and reproduction. All heterotrophic organisms require sources of chemical energy (i.e., food), water, and various limiting nutrients to survive, grow, and reproduce [@harvey_primate_1981; @baldwin_nutritional_1984; @reich_body_2001]. Failing to acquire sufficient resources is likely to result in lower fitness, changes in physiology or behavior, physiological damage, inability to reproduce, and death. Motile organisms can move to new locations in search of resources, but they must also account for a higher metabolism and movement costs [@harvey_primate_1981; @ref]. The size of an animal's spatial needs is often hypothesized to be depend on resource abundance [@burt_territoriality_1943], such that spatial needs increase when resources are low, but the relationship is likely not linear, as illustrated in figure \@ref(fig:mu-hyp) [@nilsen_can_2005; @herfindal_prey_2005; @bista_effect_2022]. Low values of $\e(R)$ force animals to have larger home ranges [@relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005; @bista_effect_2022], migrate to better locations [@samarra_movements_2017; @middleton_green-wave_2018; @geremia_migrating_2019], or move nomadically [@singh_migration_2012; @polansky_elucidating_2015; @teitelbaum_beyond_2019; @nandintsetseg_variability_2019], although large-scale changes in behavior (such as species-wide shifts to dispersal, migration, and nomadism) are more likely to occur over evolutionary timescales rather than an animal's lifespan (figure \@ref(fig:mu-hyp)). As $\e(R)$ increases, we expect animals' spatial requirements to decrease nonlinearly until they reach the smallest amount of space required to survive [see @relyea_home_2000; @bista_effect_2022, although the latter uses models which assumes the effect to be linear].

@@

Consider the simplistic scenario where the average in resource abundance, $\e(R) = \mu$, changes over time but $\var(R) = \sigma^2$ is constant over time and space (and non-zero, see figure \@ref(fig:multipanel-constant-var)a). We can use the notation $R \sim \Gamma\big(\mu(t), \sigma^2\big)$ to indicate that $R$ follows a Gamma distribution with a mean $\mu(t)$ that changes over time (i.e., it is a function of time) while the variance is constant (figure \@ref(fig:multipanel-constant-var)b). Since $R$ is spatiotemporally random, an animal that moves in the landscape will not find the same $R$ at different time points or throughout the habitat (figures \@ref(fig:multipanel-constant-var)c-d). However, if $\mu(t)$ changes repetitively and regularly over time (e.g. peaks each spring and decreases in winter), an animal may learn to predict times of high or low $R$ [e.g., @falcon-cortes_hierarchical_2021; @geremia_migrating_2019; @abrahms_memory_2019; @samarra_movements_2017; @abrahms_memory_2019] although it will not be possible to predict where high-$R$ locations are, since $\e(R)$ does not vary predictably over space.

@@

* @berger_climate_2018 fitness decreases with resources scarcity and lower temperatures

@@

Environmental variability, including extreme events [@logares_black_2012; @anderson_black-swan_2017], can reduce a landscape's energetic balance [@chevin_adaptation_2010], which, in turn, decreases animals' fitness [@berger_climate_2018] and increases their spatial needs [@refs]. A few recent studies support this hypothesis [@morellet_seasonality_2013; @nandintsetseg_variability_2019; @riotte-lambert_environmental_2020], but many of them are limited in geographic and taxonomic scales, so the extent to which these preliminary findings can be generalized is still very limited. There thus remains a need for developing a more complete understanding of how animals' spatial needs change with environmental stochasticity.

@@

Most habitats with appreciable seasonal changes in climate will require models to account for changes in both $\e(R)$ and $\var(R)$, since periods of high productivity can often be more variable than times of dormancy or lower productivity, and the variability in $R$ can often change both within and between years. In highly variable and stochastic environments, animals will likely have higher spatial needs, particularly if they are required to collect and store resources for periods of scarcity or increased energetic needs. These include reproductive periods, whether this be due to pregnancy [@ref], to feed offspring [@ref], or because the animal's needs change over time [e.g., male deer growing antlers: @french_nutrient_1956; @smolko_soil_2022]. Thus, failing to account for changes in $\var(R)$ can result in serious underestimations of animals' spatial needs, which may lead to decreases in population fitness and size while also increasing the risk of population collapse and human-wildlife conflict [@examples].

Recent increases in the frequency and intensity of extreme events [@logares_black_2012; @anderson_black-swan_2017] such as heat domes, blizzards, fires, disease outbreaks, and oil spills are important events which are hard to predict (i.e., highly stochastic) and can cause great population instability by causing mass die-offs that drastically reduce the size and genetic diversity of a population [e.g., @matkin_ongoing_2008; @murray_pathogens_2006; @widmer_effects_2004; @reed_frequency_2003; @hollings_trophic_2014; @ameca_y_juarez_natural_2012; @berger_climate_2018]. When possible, increasing the frequency of extreme events to reduce their frequency (such as with controlled burns) can increase the stability of a habitat while promoting habitat diversity and adaptations to the event [@staver_tree_2011; @tingley_pyrodiversity_2016; @nimmo_animal_2019].

@@

Not all 25 scenarios depicted in figure \@ref(fig:5-5-sims) may be realistic, but the trends in $\e(R)$ and $\var(R)$, and their impacts on animal space use are useful examples that can be thought of as simplified scenarios. $\e(R)$ and $\var(R)$ can be assumed to be (approximately) constant in highly homogeneous environments, or environments where resources are sufficiently abundant that changes in $\e(R)$ and $\var(R)$ are relatively small and remain undetected. Although it is impossible for $\e(R)$ and $\var(R)$ to increase linearly forever, such increases may be possible for short periods of time (followed by periods of no change or decrease). Cyclical oscillations in $\e(R)$ and $\var(R)$ may occur in urban environments [@peron_periodic_2017; @ikeda_home_2022] and as temperatures fluctuate daily and seasonally [@geremia_migrating_2019; @alston_temperature_2020], while $\e(R)$ and $\var(R)$ may drift randomly in highly complex environments which are too hard to predict. Finally, erratic changes in $\e(R)$ and $\var(R)$ may occur in environments where changes are very sudden, such as areas prone to fires or floods, as well as habitats with drastic human alteration (e.g., a forest which is clear-cut for mining purposes with a subsequent artificial re-forestation). However, if changes are stochastic but sufficiently small and frequent, animals may perceive them as few smooth changes rather a series of small, sudden, changes. Estimating trends in $\e(R)$ and $\var(R)$ requires a high cognitive capacity, since distinguishing between $\var (R)$ and changes in $\e(R)$ is not easy [@steixner-kumar_strategies_2020], especially if $\e(R)$ and $\var(R)$ are not independent (as in the case of $R \sim \Gamma(\mu(t), \sigma^2(t))$).

Although the trends in figure \@ref(fig:5-5-sims) are complex and the effects of $\e(R)$ and $\var(R)$ do not seem easy to disentangle, the both $\e(R)$ and $\var(R)$ affect spatial needs as we hypothesized (figure \@ref(fig:5-5-reg)). In the following discussion section, we detail how this framework can be applied to empirical data and used for conservation decision-making.

@@

This paper presents a new unifying framework for estimating the effects of resource abundance and stochasticity on animals' spatial needs. The hypotheses and methods we present allow researchers to generalize the effects resource abundance and stochasticity have on animals' spatial needs by building upon previous work rather than stating their results as *de novo* findings. Instead, one can test whether members of the same population or species behave similarly, and whether all members of the group have a common functional response [@ref]. Individuals and populations that were once thought to have markedly different behaviors and spatial needs may instead have a single common functional response but different levels of $\e(R)$ and $\var(R)$ [@examples].

In this discussion section, we make considerations about the effects of $\e(R)$ and $\var(R)$ on various animal adaptations, including movement behaviors, energy storage, and spatiotemporal memory. We then illustrate how one may apply this framework to their own data and we also provide an empirical example using tracking data on a lowland tapir (*Tapirus terrestris*) from the Brazilian Cerrado. Finally, we demonstrate how to use this framework to create products that can inform conservation-based decisions, including predicting animals' spatial needs and assessing the quality of potential (and current) conservation areas.

## Animal adaptations to changes in $\e(R)$ and $\var(R)$

@liow_lower_2009: *Lower Extinction Risk in Sleep-or-Hide Mammals*

The spatiotemporal scales over which an event occurs are a main determinant of whether an animal will be able to predict the event and how it will respond to it (or fail to do so). Events which occur at the scale of the organism [i.e. approximately one or more times per lifetime and within the animal's spatial range; see @frankham_importance_2004] are more likely to be perceived as a threat or boon worth preparing for [e.g., elephants: @foley_severe_2008; bison and mule deer: @geremia_migrating_2019].

Smaller and short-lived organisms (e.g., mice) are more likely to be be severely impacted by stochastic events than larger ones (e.g. elephants), since larger organisms can have bigger energy reserves [@lindstedt_seasonality_1985], can move longer distances over short periods of time [@hirt_general_2017], and tend to have longer lifespans, generation times, and developmental periods [@brown_toward_2004] that allow them to develop or memory about the frequency and severity of such events [@foley_severe_2008; @polansky_elucidating_2015]. However, the short generation time and high reproductive rate of smaller *r-selected* species [@pianka_r-_1970; @brown_toward_2004] may also promote traits that increase survival following frequent extreme events by increasing population survival rates and genetic diversity through the production of numerous offspring [@examples]. Frequent events may result in changes in animal behavior and memory [framework: @riotte-lambert_environmental_2020; elephants: @foley_severe_2008; bison and mule deer: @geremia_migrating_2019; elk: @falcon-cortes_hierarchical_2021], while events that occur over larger spatiotemporal scales are more likely to promote slower changes through adaptation, natural selection, and evolution [@grant_evolution_2017; @logares_black_2012; @anderson_black-swan_2017; @gienapp_climate_2008; @examples], including the developing of memory [framework: @riotte-lambert_environmental_2020; @fagan_spatial_2013; roe deer: @ranc_memory_2022; elk: @falcon-cortes_hierarchical_2021; blue whales: @abrahms_memory_2019; elephants: @polansky_elucidating_2015].

Adaptability and responsiveness to change are strong determinants to whether an animal survives in a changing environment [@riotte-lambert_environmental_2020]. An animal's ability to recognize cues that precede extreme events or periods of scarcity or abundance can have appreciable effects on the animal's chance of survival and reproduction. For example, ungulates' ability to time their yearly migrations on cues such as available forage biomass, hunting pressure, snow fall, and snow melt allow them to adapt to changes in both $\e(R)$ and $\var(R)$, while their spatial memory [@falcon-cortes_hierarchical_2021; @ranc_memory_2022] allows them to optimize their movement by reducing the amount of time spent exploring. Similarly, the ability of trans-Saharian migrant birds to adapt their migration timing based on environmental cues allows them to account for differences $\e(R)$ between years [@jonzen_rapid_2006]. However, relying strongly on short-term cues rather than depending on memory can also cause animals to fail to find crucial resources [@polansky_elucidating_2015; @foley_severe_2008] or become trapped in sub-optimal locations if long-term conditions and cycles change unexpectedly [e.g., blue whales: @abrahms_memory_2019; bison: @merkle_spatial_2019].

@@

The hypotheses we present here allow researchers to combine findings and generalize results to a single, common set of functional responses. We expect the two hypotheses we present here (figures \@ref(fig:mu-hyp) and \@ref(fig:s2-hyp)) to be applicable to all animals, once differences in metabolic needs and diet are accounted for. In this subsection, we illustrate how researchers can test these hypotheses using their own data to build on previous work rather than listing the results as *de novo* findings. Additionally, the analyses can be expanded to test for commonalities and differences in the behaviors between members of the same population or species, as well as between multiple distinct populations or species.

Ecologists often conclude that populations from different regions behave differently [@examples], but the differences in behavior may only be due to differences in $\e(r)$ or $\var(R)$, instead, which would cause animals to behave differently even if they depend on the same functional responses.

We start by providing examples and considerations on how one may model $R$ to estimate $\e(R)$ and $\var(R)$, animals' spatial needs, and the effects of $\e(R)$ and $\var(R)$ on animal's spatial needs. Next we provide an empirical example using tracking data on a lowland tapir (*Tapirus terrestris*) from the Brazilian Cerrado and NDVI as a proxy for $R$. Finally, we offer suggestions on how this approach can be used to inform conservation-related decisions, including assessing habitat quality and estimating animals spatial needs under different scenarios.

@@

Location-scale models are a class of statistical models that allow one to model the mean and variance of the response rather than only modeling the mean.

@@

Any linear model, including ANOVA-like models, would assume spatial needs can span all (complex) numbers, including negative values. Since spatial needs are strictly greater than zero, models that allow non-linear relationships while forcing the spatial needs to be strictly greater than zero.

@@

### An empirical example: Lowland tapir in the Brazilian Cerrado

@@

## Using this framework in conservation-related decisions
